Exploiting First-Class Arrays in Fortran for Accelerator Programming

### 2010-12-16 opencl/papers/intro.tex
LA-UR version

This paper presents a compiler-level approach for targeting a single
program to multiple fundamentally different low level execution and
programming models while allowing the application programmer to adopt
a single high level programming model.  We show that features of
Fortran 90 for data parallel programming are well suited to automatic
transformation to generate code specifically tuned for different
low-level programming models such as OpenCL and CUDA.  For algorithms
that can be easily expressed in terms of whole array, data parallel
operations, writing code in Fortran and transforming it automatically
to specific low level implementations removes the burden from the
programmer of working with tedious, error prone, low-level tools.

In the ideal situation, application programmers would like to adopt
a programming model in which they write their application once and
use automated tools to retarget it to many architectures.  This has
proven to be very challenging historically due to the subtle balance
between high-level expressiveness of code and the performance of the
lower-level code that is emitted by a compiler.  This ideal high-level
model that programmers work with should emphasize readability,
maintainability, and close proximity in abstraction to the problem being
solved.  It should not be corrupted with details of specific target
architectures solely for the purpose of single-system performance.
For certain classes of applications, specifically those that map well
onto a data-parallel programming model, we show that Fortran 90 contains
language features that encourage high-level programming abstractions
without sacrificing performance during low-level code generation.

@@

At Los Alamos National Laboratory (LANL), as with many supercomputing
facilities today, users have a wide variety of computer platforms from
which to choose.  The most common platform is made up of clusters of
compute nodes with standard multi-core processors.  An increasingly
common feature is that some nodes also have accelerators that range
from the IBM Cell processor (such as the LANL Roadrunner system) to a
variety of GPUs from NVIDIA and AMD.  Some nodes have hardware with
vector instructions and others do not.  The peak performance of these
accelerated nodes often resides in the hundreds of gigaflops.

@@

A common theme amongst new processors is the emphasis on data parallel
programming.  This model is well suited to emerging architectures that
are based on either vector processing or massively parallel
collections of simple cores.  A popular trend for programming systems
such as GPU accelerators is the CUDA language from NVidia, or new
languages like Chapel from Cray.  A less intrusive approach that is
well suited to legacy applications and languages are language
extensions or libraries that allow programmers to avoid adopting
entirely new languages.  The recent OpenCL specification is intended
to support this programming model, as are directive-based methods such
as OpenMP or the Accelerator programming model from the Portland
Group~\cite{pgi10accelerator}.

@@

At this point in time it is not really possible to write once and run
efficiently on the wide variety of computer platforms we have
available.  For some classes of applications, we believe that this
goal is possible using language constructs already present in a
popular mainstream scientific programming language -- Fortran.  A
common, long-standing tongue-in-cheek response to new language
developments in the scientific and high performance computing
community is that a new language will arise to answer the needs of new
systems, and it will be called Fortran.  We believe that the work
presented in this paper validates that notion -- we need a new
language to work with, and that language is Fortran.

@@

This paper addresses the accelerator programmign problem by examining
features in Fortran that allow programmers to express algorithms at a
very high level that can be easily transformed by a compiler to run
efficiently on a wide variety of platforms.  In particular we consider
computers based on GPUs and related accelerator processors.

We demonstrate that the array syntax of Fortran maps surprisingly well
onto GPUs when transformed to OpenCL kernels.  These Fortran language
features include pure and elemental functions and array constructs
like where and shift.  In addition we add a few functions that enable
a program to be deployed on machines with a hierarchy of processing
elements, such as nodes employing GPU acceleration, \emph{without
  requiring explicit declaration of parallelism within the program.}
In addition the program uses entirely standard Fortran so it can be on
a single core without concurrency.  This work also is applicable
to vendor-specific languages similar to OpenCL such as the NVidia
CUDA language.

We provide (via Fortran interfaces in the ForOpenCL library) a mechanism to
call the C OpenCL runtime and enable Fortran programmers to access OpenCL
kernels.  Transformations are supplied that provide a mechanism for converting
Fortran procedures written in the Fortran subset described in this paper to
OpenCL kernels.  We use the ROSE compiler infrastructure \cite{} to develop
these transformations.  ROSE uses the Open Fortran Parser \cite{} to parse
Fortran 2008 syntax and can output (unparse) to OpenCL.  Since ROSE's
intermediate representation (IR) was constructed to represent multiple
languages, it is relatively straightforward to transform high-level Fortran IR
nodes to OpenCL nodes.  Furthermore, the Fortran array syntax maps directly to
one, two, and three-dimensional thread groups in OpenCL.

@@

We study automatic transformations and the performance for an
application example that is typical of many applications that are based on
finite-difference or finite-volume methods in computational fluid dynamics
(CFD).  The example is a simple shallow water model in two dimensions using
finite volume methods with .... for time update.

An initial study was made for an important procedure in PAGOSA, a
non-research, production-grade code at LANL completely written in
data-parallel Fortran.  We investigated automatically transforming
this code to run on LANL's Petaflop Roadrunner computer (a hybrid
mixture of AMD Opterons and IBM Cell processors).  We demonstrated that a
source-to-source compiler can automatically vectorize and parallelize
(both within the Cell processor and across nodes) a small section of
this code.  Preliminary results showed a 12 times performance gain on
the Cell processor over a traditional single core processor.  A factor
of 9 is gained from vectorizing and parallelizing the code to run on
the Cell and the rest is gained from automatically overlapping
inter-node communication with computation on the Cell.

@@

Fortran 
%(originally named FORTRAN, the FORmula TRANslation
%language\footnote{The capitalization convention for referring the
%  language was changed in 1990 to \emph{Fortran}}) 
is the oldest
high-level programming language in continuous use since its
introduction, and was developed to facilitate the translation of math
formulae into machine code. Fortran was the first major language to
use a compiler to translate from a high-level program representation
to assembly language. Due to its age, it carries certain arcane
baggage that later programming languages have evolved away from. As a
result, Fortran has fallen into disfavor in certain programming
circles.  Modern versions of the language standardized in 1990, 1995, 2003, and 2008
have removed much of this legacy baggage, but these changes are not
widely known.  Modern Fortran exhibits features that are similar to 
other modern programming languages, and does not mandate legacy features
from decades old, deprecated versions of the language.

With the introduction of Fortran 90 (and later revisions to the
standard), Fortran became a truly modern programming language. Much of
the arcane baggage associated with it is actually no longer part of
the language used by modern programmers, and exists in a deprecated
form simply to support legacy codes. It is now modular and has many
object-oriented features.  It now includes a type system in which rich
first-class array data types and corresponding syntax are part of the
language, something that languages like C continue to lack.
Furthermore, Fortran should be of interest to those studying parallel
programming because of its functional and data parallel constructs and
because of the coarray notation introduced in Fortran 2008. Unlike
languages like C and C++, which are considered to be more modern,
Fortran has become a truly parallel language with features added to
recent language standards.

Yet, a likely question that one may pose is ``\emph{Why Fortran and not a more
  modern language like X?}''  The recent rise in interest in concurrency and
parallelism at the language level driven by multicore CPUs and manycore
accelerators has driven a number of new language developments, both as novel
languages and extensions on existing ones.

%% NEW - CER
So perhaps it is time to replace Fortran with yet another computer language.  The
problems with replacing Fortran with an entirely new language are two fold:
the economics of replacing the existing application base and the difficulty in
obtaining programmer acceptance.  It is estimated that replacing a major
production application at Los Alamos National Laboratory would cost between 50
and 150 million dollars.  In terms of programmer acceptance, there is always
the "chicken and egg problem": programmers won't use a new language until they
can expect good performance across a variety of platforms, and compiler
vendors can't afford to produce quality compilers until there is a reasonable
expectation of a market.

For scientific users, these new languages and language extensions
present a challenge: how do developers effectively use them while
avoiding rewriting their code and potentially growing dependent on a
transient technology that will vanish tomorrow?

%% NEW - CER
History has also shown that an investment in rewriting code does not guarantee
success either, as seen in an effort at LANL to modernize a legacy Fortran
code with the the newer C++ POOMA framework.  This massive overhaul effort led
to a code that was both slower and less flexible than the original Fortran
\cite{}.  Similar experiences have occurred in the past, notably during the
devlopment of the Sisal language.

@@

The High Performance Fortran (HPF) extension of Fortran 90 was
proposed to add features to the language that would enhance the
ability of compilers to emit fast parallel code for distributed and
shared memory parallel computers\cite{koelbel94hpf}.  One of the
notable additions to the language in HPF was syntax to specify the
distribution of data structures amongst a set of parallel processors.
HPF also introduced an alternative looping construct to the
traditional DO-loop called \emph{FORALL} that was better suited for
parallel compilation.  An additional keyword, \emph{INDEPENDENT}, was
added to allow the programmer to indicate when the order of execution
of the program (such as a sequence of loop iterations) can be flexible
in order to allow parallel execution.  Interestingly, the parallelism
features introduced in HPF did not exploit the new array features
introduced in 1990 in any significant way, relying instead on explicit
loop-based parallelism.  This was likely in order to support parallel
programming that wasn't easily mapped onto a pure data parallel model.

@@

When shifts are employed, this can be expressed as:

@@

\begin{verbatim}
Write some ZPL
\end{verbatim}

This would then be executed over the entire array. {\bf more...}

Stuff on stencil compilers?  No?

>>>>

This paper presents a compiler-level approach for targeting a single
program to multiple fundamentally different low-level execution and
programming models while allowing the application programmer to adopt
a single high-level programming model.  We show that features of
Fortran 90 for data parallel programming are well suited to automatic
transformation to generate code specifically tuned for different
low-level programming models such as OpenCL and CUDA.  For algorithms
that can be easily expressed in terms of whole array, data parallel
operations, writing code in Fortran and transforming it automatically
to specific low-level implementations removes the burden from the
programmer of working with tedious, error prone, low-level tools.

In the ideal situation, application programmers would like to adopt a
programming model in which they write their application once and use
automated tools to retarget it to many architectures.  This has proven
to be very challenging historically due to the subtle balance between
high-level expressiveness of code and the performance of the
lower-level code that is emitted by a compiler.  This ideal high-level
model that programmers work with should emphasize readability,
maintainability, and close proximity in abstraction to the problem
being solved --- in this instance, the abstraction that we care about
are mathematical formulae.  It should not be corrupted with details of
specific target architectures solely for the purpose of single-system
performance.  For certain classes of applications, specifically those
that map well onto a data-parallel programming model, we show that
Fortran 90 contains language features that encourage high-level
programming abstractions without sacrificing performance during
low-level code generation.

@@

At the Los Alamos National Laboratory (LANL), as with many supercomputing
facilities today, users have a wide variety of computer platforms from
which to choose.  The most common platform is made up of clusters of
compute nodes with standard multi-core processors.  An increasingly
common feature is that some nodes also have accelerators that range
from the IBM Cell processor (such as the LANL Roadrunner system) to a
variety of GPUs from NVIDIA and AMD.  Some nodes have hardware with
vector instructions and others do not.  The peak performance of these
accelerated nodes often resides in the hundreds of gigaflops.

@@

A common theme amongst new processors is the emphasis on data parallel
programming.  This model is well suited to emerging architectures that
are based on either vector processing or massively parallel
collections of simple cores.  A proposed approach for programming that
is well suited to legacy applications and languages are language
extensions or libraries that allow programmers to avoid adopting
entirely new languages.  The recent OpenCL specification is intended
to support this programming model, as are directive-based methods such
as OpenMP or the Accelerator programming model from the Portland
Group~\cite{pgi10accelerator}.

@@

%At this point in time it is not really possible to write once and run
%efficiently on the wide variety of computer platforms we have
%available.  For some classes of applications, we believe that this
%goal is possible using language constructs already present in a
%popular mainstream scientific programming language -- Fortran.  A
%common, long-standing tongue-in-cheek response to new language
%developments in the scientific and high performance computing
%community is that a new language will arise to answer the needs of new
%systems, and it will be called Fortran.  We believe that the work
%presented in this paper validates that notion -- we need a new
%language to work with, and that language is Fortran.

@@

%This paper addresses the accelerator programmign problem by examining
%features in Fortran that allow programmers to express algorithms at a
%very high level that can be easily transformed by a compiler to run
%efficiently on a wide variety of platforms.  In particular we consider
%computers based on GPUs and related accelerator processors.

We demonstrate that the array syntax of Fortran maps surprisingly well
onto GPUs when transformed to OpenCL kernels.  These Fortran language
features include pure and elemental functions and array constructs
like {\tt where} and {\tt cshift}.  In addition we add a few functions
that enable a program to be deployed on machines with a hierarchy of
processing elements, such as nodes employing GPU acceleration,
\emph{without requiring explicit declaration of parallelism within the
  program.}  In addition the program uses entirely standard Fortran so
it can be compiled for and executed on a single core without concurrency.
This work also is applicable to vendor-specific languages similar to
OpenCL such as the NVidia CUDA language.

%We provide (via Fortran interfaces in the ForOpenCL library) a
%mechanism to call the C OpenCL runtime and enable Fortran programmers
%to access OpenCL kernels.  
Transformations are supplied that provide a mechanism for converting
Fortran procedures written in the Fortran subset described in this
paper to OpenCL kernels.  We use the ROSE compiler infrastructure
\cite{rose-web} to develop these transformations.  ROSE uses the Open
Fortran Parser \cite{ofp:sf} to parse Fortran 2008 syntax and can
generate C-based OpenCL.  Since ROSE's intermediate representation
(IR) was constructed to represent multiple languages, it is relatively
straightforward to transform high-level Fortran IR nodes to C OpenCL
nodes.

%  Furthermore, the Fortran array syntax maps directly to one,
%two, and three-dimensional thread groups in OpenCL.

@@

We study automatic transformations and the performance for an
application example that is typical of many applications that are
based on finite-difference or finite-volume methods in computational
fluid dynamics (CFD).  The example described later in this paper is a
simple shallow water model in two dimensions using finite volume
methods.

An initial study was made for an important procedure in PAGOSA, a
non-research, production-grade code at LANL completely written in
data-parallel Fortran.  We investigated automatically transforming
this code to run on LANL's Petaflop Roadrunner computer (a hybrid
mixture of AMD Opterons and IBM Cell processors).  We demonstrated that a
source-to-source compiler can automatically vectorize and parallelize
(both within the Cell processor and across nodes) a small section of
this code.  Preliminary results showed a 12 times performance gain on
the Cell processor over a traditional single core processor.  A factor
of 9 is gained from vectorizing and parallelizing the code to run on
the Cell and the rest is gained from automatically overlapping
inter-node communication MPI with computation on the Cell.

@@

Fortran is the oldest high-level programming language in continuous
use since its introduction, and was developed to facilitate the
translation of math formulae into machine code. Fortran was the first
major language to use a compiler to translate from a high-level
program representation to assembly language. Due to its age, it
carries certain arcane baggage that later programming languages have
evolved away from. As a result, Fortran has fallen into disfavor in
certain programming circles.  Modern versions of the language
standardized in 1990, 1995, 2003, and 2008 have removed much of this
legacy baggage, but these changes are not widely known.  Modern
Fortran exhibits features that are similar to other modern programming
languages, and does not mandate the use of legacy features from
decades old, deprecated versions of the language.  Compilers for
Fortran (like those for other languages that have removed features
over time) can prohibit the use of archaic features such as fixed
format code or constructs that have been removed from the language.

With the introduction of Fortran 90 (and later revisions to the
standard), Fortran became a truly modern programming language. Much of
the arcane baggage associated with it is actually no longer part of
the language used by modern programmers, and exists in a deprecated
form simply to support legacy codes. It is now modular and has many
object-oriented features.  Most importantly for this work, it now
includes a type system in which rich first-class array data types and
corresponding syntax are part of the language, something that
languages like C continue to lack.  Furthermore, Fortran should be of
interest to those studying parallel programming because of its
functional and data parallel constructs and because of the coarray
notation introduced in Fortran 2008. Unlike languages like C and C++,
which are considered to be more modern, Fortran has become a truly
parallel language with features added to recent language standards.

Yet, a likely question that one may pose is ``\emph{Why Fortran and
  not a more modern language like X?}''  The recent rise in interest
in concurrency and parallelism at the language level driven by
multicore CPUs and manycore accelerators has driven a number of new
language developments, both as novel languages and extensions on
existing ones.  For scientific users, new languages and language
extensions to use novel new architectures present a challenge: how do
developers effectively use them while avoiding rewriting their code
and potentially growing dependent on a transient technology that will
vanish tomorrow?

%% NEW - CER
%So perhaps it is time to replace Fortran with yet another computer language.  The
%problems with replacing Fortran with an entirely new language are two fold:
%the economics of replacing the existing application base and the difficulty in
%obtaining programmer acceptance.  It is estimated that replacing a major
%production application at Los Alamos National Laboratory would cost between 50
%and 150 million dollars.  In terms of programmer acceptance, there is always
%the "chicken and egg problem": programmers won't use a new language until they
%can expect good performance across a variety of platforms, and compiler
%vendors can't afford to produce quality compilers until there is a reasonable
%expectation of a market.

%% NEW - CER
History has also shown that an investment in rewriting code does not guarantee
success either, as seen in an effort at LANL to modernize a legacy Fortran
code with the the newer C++ POOMA framework.  This massive overhaul effort led
to a code that was both slower and less flexible than the original Fortran
\cite{basili08hpc}.  Similar experiences have occurred in the past, notably during the
devlopment of the Sisal language.

@@

The High Performance Fortran (HPF) extension of Fortran 90 was
proposed to add features to the language that would enhance the
ability of compilers to emit fast parallel code for distributed and
shared memory parallel computers\cite{koelbel94hpf}.  One of the
notable additions to the language in HPF was syntax to specify the
distribution of data structures amongst a set of parallel processors.
HPF also introduced an alternative looping construct to the
traditional DO-loop called {\tt FORALL} that was better suited for
parallel compilation.  An additional keyword, {\tt INDEPENDENT}, was
added to allow the programmer to indicate when the order of execution
of the program (such as a sequence of loop iterations) can be flexible
in order to allow parallel execution.  Interestingly, the parallelism
features introduced in HPF did not exploit the new array features
introduced in 1990 in any significant way, relying instead on explicit
loop-based parallelism.  This was likely in order to support parallel
programming that wasn't easily mapped onto a pure data parallel model.

@@

\noindent When shifts are employed, this can be expressed as:

@@

%%TODO
%%\begin{verbatim}
%%Write some ZPL
%%\end{verbatim}
%%
%%This would then be executed over the entire array. {\bf more...}
%%
%%Stuff on stencil compilers?  No?

### 2010-12-16 opencl/papers/programming-model.tex

The static analysis and source-to-source transformations used are very
basic and only require the programmer to use a subset of Fortran that
employs a data-parallel programming model.  In particular, it
encourages use of language features that were introduced and
standardized in the Fortran 90 language specification.  In this
section we describe the set of four Fortran 90 features that our analysis
and transformation method are based on.  From these language constructs,
we are able to easily transform them to a lower-level CUDA or OpenCL
implementation.

@@

Fortran 90 introduced a rich array syntax that allows programmers to
write statements that are in terms of whole arrays or subarrays, with
data parallel operators to compute on the arrays.  This has the
benefit of avoiding explicit looping in the code and maintaining a
high-level style that is closer to the original mathematical
specification of the problem being solved.  More importantly from a
compilation perspective, this defers decisions about how to implement
these whole-array operations to a compilation tool.  Sophisticated
loop analysis to identify parallelism within the loop are not
necessary.  When faced with a novel architecture such as modern GPUs
that are ideally suited to data parallel programming models, the fit
between Fortran arrays and these systems is quite clean.

%% NEW - CER
Arrays are first class citizens in Fortran instead of a basic pointer
to raw memory as in many other compiled languages.  A Fortran array
contains metadata associated with it that describes the array's shape
and size. It should be noted that other languages that provide rich
array data types and whole array operations may also be suitable
targets for the transformations described in this paper.  
In addition to metadata about the structure of the arrays, operators
are provided that allow programmers to write expressions based on the
whole array instead of explicit indexing and application of operations
to individual elements.  For example, if {\tt A}, {\tt B}, and {\tt C}
are all arrays of the same rank and shape and {\tt s} is a scalar,
then the statement

@@

results in the element-wise sum of {\tt A} and {\tt s} times the elements of
{\tt B} being stored in the corresponding elements of {\tt C}. The first
element of {\tt C} will contain the value of the first element of {\tt A}
added to the first element of {\tt c*B}.  Note that no explicit iteration over
array indices is needed and that the individual operators, plus, times, and
assignment are applied by the compiler to individual elements of the arrays
independently.  Thus the compiler is able to spread the computation in the
example across any hardware threads under its control.

@@

An elemental function consumes and produces scalar values, but can be applied
to variables of array type such that the function is applied to each and every
element of the array.  This allows programmers to avoid explicit looping and
instead simply state that they intend a function to be applied to every
element of an array in parallel, deferring the choice of implementation
technique to the compiler.  Elemental functions are intended to be used for
data parallel programming, and as a result must be side effect free (or,
\emph{pure}).

@@

Note that while {\tt foo} is defined in terms of purely scalar
quantities, it can be \emph{applied} to arrays as shown.  While this
may seem like a trivial example, such simple functions may be composed
with other elemental functions to perform powerful computations,
especially when applied to arrays.  Our prototype tool transforms
elemental functions to inline OpenCL functions.  Thus there is no
penalty for usage of elemental functions and provide a convient
mechanism to express algorithms in simpler segments.

@@

%% NEW - CER
Complete and very concise and elegant programs can be built with procedures
similar to the example shown above. To aid this effort, Fortran
supplies intrinsic functions like the array constructors (\texttt{CSHIFT},
\texttt{EOSHIFT}, \texttt{MERGE}, \texttt{TRANSPOSE}, ...), the array location
functions (\texttt{MAXLOC} and \texttt{MINLOC}), and the array reduction
functions (\texttt{ANY}, \texttt{COUNT}, \texttt{MINVAL}, \texttt{SUM},
\texttt{PRODUCT}, ...).  To this set we add region functions described above.

>>>>

The static analysis and source-to-source transformations used are very
basic and only require the programmer to use a subset of Fortran that
employs a data-parallel programming model.  In particular, it
encourages use of four language features introduced in Fortran 90 ---
array notation, shift functions, elemental functions, and pure
procedures.  From these language constructs, we are able to easily
transform them to a lower-level CUDA or OpenCL implementation.

@@

Fortran 90 introduced a rich array syntax that allows programmers to
write statements that are in terms of whole arrays or subarrays, with
data parallel operators to compute on the arrays.  This has the
benefit of avoiding explicit looping in the code and maintaining a
high-level style that is closer to the original mathematical
specification of the problem being solved.  More importantly from a
compilation perspective, this defers decisions about how to implement
these whole-array operations to a compilation tool.  Sophisticated
loop analysis to identify parallelism within the sequential loop code
are not necessary.  When faced with a novel architecture such as
modern GPUs that are ideally suited to data parallel programming
models, the fit between Fortran arrays and these systems is quite
clean.

A Fortran array combines a base pointer to memory with metadata that
describes the array shape, stride and size. It should be noted that
other languages that provide rich array data types and whole array
operations may also be suitable targets for the transformations
described in this paper.  Array variables can be used in expressions
based on whole-array operations.  For example, if {\tt A}, {\tt B},
and {\tt C} are all arrays of the same rank and shape and {\tt s} is a
scalar, then the statement

@@

\noindent results in the element-wise sum of {\tt A} and {\tt s} times
the elements of {\tt B} being stored in the corresponding elements of
{\tt C}. The first element of {\tt C} will contain the value of the
first element of {\tt A} added to the first element of {\tt c*B}.
Note that no explicit iteration over array indices is needed and that
the individual operators, plus, times, and assignment are applied by
the compiler to individual elements of the arrays independently.  Thus
the compiler is able to spread the computation in the example across
any hardware threads under its control.

@@

An elemental function consumes and produces scalar values, but can be applied
to variables of array type such that the function is applied to each and every
element of the array.  This allows programmers to avoid explicit looping and
instead simply state that they intend a function to be applied to every
element of an array in parallel, deferring the choice of implementation
technique to the compiler.  Elemental functions are intended to be used for
data parallel programming, and as a result must be side effect free and
mandate an {\tt intent(in)} attribute for all arguments.

@@

Note that while {\tt foo} is defined in terms of purely scalar
quantities, it can be \emph{applied} to arrays as shown.  While this
may seem like a trivial example, such simple functions may be composed
with other elemental functions to perform powerful computations,
especially when applied to arrays.  Our prototype tool transforms
elemental functions to inline OpenCL functions.  Thus there is no
penalty for usage of elemental functions and they provide a convient
mechanism to express algorithms in simpler segments.

@@

%% NEW - CER
%Complete and very concise and elegant programs can be built with procedures
%similar to the example shown above. To aid this effort, Fortran
%supplies intrinsic functions like the array constructors (\texttt{CSHIFT},
%\texttt{EOSHIFT}, \texttt{MERGE}, \texttt{TRANSPOSE}, ...), the array location
%functions (\texttt{MAXLOC} and \texttt{MINLOC}), and the array reduction
%functions (\texttt{ANY}, \texttt{COUNT}, \texttt{MINVAL}, \texttt{SUM},
%\texttt{PRODUCT}, ...).  To this set we add region functions described above.

### 2010-12-16 opencl/papers/source-to-source.tex

OpenCL~\cite{opencl:standard} is an open language standard for developing
applications for accelerators.  The C-based language provides extensions for
programming kernels that run on accelerator processing elements.  The kernels
are run by calling a C runtime library from the OpenCL host (normally the
CPU).  Efforts to standardize a C++ runtime are underway and Fortran
interfaces to the C runtime are described later.

@@

This work describes transformations that automatically create OpenCL kernel
functions from Fortran pure and elemental procedures.  These transformations
will not transform an entire program.  Users, for now, must explicitly replace
calls to Fortran kernel procedures that run on the device, with calls to the
OpenCL runtime that will run the kernel on the OpenCL host.  While these host
transformations are straightforward using ROSE, they are outside the scope of
this paper.

In addition to the Fortran to OpenCL transformations, the
ForOpenCL~\cite{http:ofp} library provides programmers with the ability to
call the C OpenCL runtime from Fortran.  ForOpenCL is a set of Fortran modules
providing Fortran 2003 interface descriptions and classes that allow language
interoperability with the OpenCL runtime.

@@

\subsection{Performance measurements}

Performance measurements were made comparing the transformed code with
different versions of the serial shallow-water code.  The serial versions
include the original written in C and two Fortran versions: one using copy
semantics for the regions and one using reference semantics.  We also compare
with a hand written OpenCL implementation that was optimized for local memory
usage (no array temporaries).  The measurments were made using an NVIDIA Tesla C2050
GPU (TODO - need model numbers) running at XXX Hz and an Intel XXX hexacore CPU
running at XXX Hz with 2.56 GB of RAM.  The compilers were gfortran and gcc VERSIONS with
optimization of -O3.

The performance comparison is shown in Figure~\ref{fig:cl-performance} for
varying array sizes (of the state variables).  The time represents an average
of 100 iterations of the outer time-advance loop calling the OpenCL kernel.
This tight loop keeps the OpenCL kernel supplied with threads to take
advantage of potential latency hiding by the NVIDIA GPU.  Any serial code
within this loop would reduce the measured values.

As can be seen in Figure~\ref{fig:cl-performance}, the transformed get very
good results of roughly 65 times speedup over the best Fortran code and YYY
speedup over the original C code.  It comes with XXX \% of the hand coded
OpenCL version.

%\begin{figure}[!t]
%\centering
%\includegraphics[width=2.5in]{cl-performance.pdf}
%\caption{Performance comparison for varying array size.}
%\label{fig:cl-performance}
%\end{figure}

>>>>

OpenCL~\cite{opencl08} is an open language standard for developing
applications for accelerators.  The C-based language provides extensions for
programming kernels that run on accelerator processing elements.  The kernels
are run by calling a C runtime library from the OpenCL host (normally the
CPU).  Efforts to standardize a C++ runtime are underway and Fortran
interfaces to the C runtime are described later.

@@

This work describes transformations that automatically create OpenCL kernel
functions from Fortran pure and elemental procedures.  At present, these transformations
will not transform an entire program.  Users, for now, must explicitly replace
calls to Fortran kernel procedures that run on the device, with calls to the
OpenCL runtime that will run the kernel on the OpenCL host.  While these host
transformations are straightforward using ROSE, they are outside the scope of
this paper.

In addition to the Fortran to OpenCL transformations, the
ForOpenCL~\cite{ofp:sf} library provides programmers with the ability to
call the C OpenCL runtime from Fortran.  ForOpenCL is a set of Fortran modules
providing Fortran 2003 interface descriptions and classes that allow language
interoperability with the OpenCL runtime.

@@
